---
title: "Capital Punishment in the U.S."
subtitle: "STT2860 Project 02 Fall 2022"
author: "Gracey, Natara, Evan"
date: "`r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: yeti
    highlight: textmate
    toc: TRUE
    toc_depth: 4
---
<hr>

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

### Packages Used

```{r}
library(tidyverse)
```

### Read Data Files

```{r}

cappunish <- readRDS("cappunish.rds")
deathrowsize <- readRDS("deathrowsize.rds")
deathsentencesNC <- readRDS("deathsentencesNC.rds")
```

### Plot #1: How Many States Have Capital Punishment? 

* Create a barplot for the variable `dp2`. 
* Use `color =` and `fill =` to format the bars.
* Use `labs()` to add a title and better axis labels.
* Apply a built-in theme from **ggplot2** or **ggthemes**.
* You may optionally add a subtitle or caption to the plot.
* This plot should *not* have a visible legend anywhere on it.

```{r}
ggplot(cappunish, aes(x = dp2)) +
  geom_bar(color = 'white', fill = "#906388") + 
  labs(title = "States with Capital Punishment")
  
```

### Plot #2: Capital Punishment By Region of the U.S.

* Create a barplot for the variable `region`. 
* Use `fill =` to shade each bar according to `dp2`.
* Change the `position =` to set each bar equal to 100%.
* Use `width =` to make the bars narrower than the default.
* Use `labs()` to add a title, axis labels, and a better legend label.
* Use `scale_fill_manual()` to choose a different color scheme for the plot. 
* Use `theme()` to move the legend to the bottom of the plot instead of the right
* You may optionally add a subtitle, caption, or ggplot2 built-in theme to the plot.

```{r}
ggplot(cappunish, aes(x = region, fill = dp2)) +
  geom_bar() +
  labs(title = "Region acording to dp2") +
  scale_fill_manual(values = c("#D0B0AF", "#80AF90", "#8090AF"))+
  theme(legend.position="bottom")

```

### Plot #3: Number of Prisoners on Death Row

* Create a barplot for the total number of prisoners (y) in each year (x).
* Fill the bars using the variable `decade` and outline them all in black.
* Set the width of all the bars so there is no visible gap in between them.
* Scale the x-axis to include a tick mark at each decade (1970, 1980, etc).
* Scale the y axis so that it is numbered 0, 500, 1000, 1500, 2000, etc.
* Use `labs()` to add or change the title, axis labels, and legend label. 
* Use `theme()` to do the following:
    * center your title
    * set the plot background to white or blank
    * set all axis tick marks to transparent or blank
    * set the major x-axis gridlines to transparent or blank
    * set the major y-axis gridlines to light gray
* You may optionally change the bar color scheme used in your plot

```{r}
ggplot(deathrowsize, aes(x = year, y = total, fill = decade)) +
  geom_col(color = "black") +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  scale_y_continuous(breaks = seq(0, 10000, by = 500)) +
  labs(title = "Prisoners per Decade") +
  theme(panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.title.x = element_blank()) +
   theme(axis.title.y = element_text(color = "light gray"))

```

### Plot #4: Number of Executions Since 1976

* Create a violin plot for the variable `post1976` (use `y = 0`)
* Add a boxplot and set the width so it fits inside the violin plot
* Fill each plot with a color of your choice (be colorblind-aware)
* Scale the x-axis of the plot so that it counts in increments of 25
* Get rid of all y-axis markings/gridlines using `scale_y_continuous()`
* Add a title, x-axis label, and a caption to identify the large outlier
* Add the built-in theme `theme_linedraw()` or `theme_minimal` to the plot

Hint: You can use the `NULL` vector in functions like `scale_y_continuous()`

```{r}
ggplot(cappunish, aes(y = 0, x = post1976)) +
geom_violin(fill = "#56B4E9")  +
  geom_boxplot(fill = "#D55E00", width = .1) +
   scale_x_continuous(breaks = seq(0,2000, by = 25)) +
    scale_y_continuous(breaks = NULL) +
  labs(title = "Executions Post 1976", x = "Number of Executions", caption = "One state, Texas, has conducted an abnormally large number of executions - around 575 - since 1976.") +
  theme_linedraw() +
  theme(axis.title.y = element_blank())
  
  
```

### Plot #5: Death Sentences in NC Since 1976

Create a time series plot of death sentences in North Carolina. Use three geometries on this plot: `geom_area()`, `geom_line()` and `geom_point()`. Use `geom_vline()` to add a vertical line for 2001. The significance of this year is that the General Assembly modified the death penalty statute in several ways, including giving prosecutors the discretion to *not* seek the death penalty in cases where it previously would have been mandatory. Format and label your plot in such a way that the information it conveys would be clear to the reader (i.e., the plot is reasonably stand-alone).

```{r}
ggplot(deathsentencesNC, aes(x = year, y = sentences)) +
  geom_area(fill = "#57C4AD") +
  geom_line(color = "#B3589a") + 
  geom_point(color = "#EDA247") + 
  geom_vline(xintercept = 2001) +
  scale_x_continuous(breaks = seq(1960,2025, by = 5)) +
labs(title = "Death Sentences in NC since 1976", y = "Number of Sentences")
```

### Information Conveyed by Your Plots

List two pieces of information you can learn or comparisons you can make from each plot above.

#### Plot #1

* Item 1: learned how many states have capital punishment
* Item 2: learned how many states have capital punishment under moratorium, meaning it temporarily stopped

#### Plot #2

* Item 1: learned what regions where there is more states with capital punishment 
* Item 2: learned what regions where capital punishment is in moratorium

#### Plot #3

* Item 1: learned that the total number of prisoners was lowest in the 70's 
* Item 2: learned that the total number of prisoners was highest at the beginning of the 2000's

#### Plot #4

* Item 1: learned how many executions there has been since 1976
* Item 2: can see which states have had the most executions in 1976

#### Plot #5

* Item 1: the number of death sentences was the highest in 1995
* Item 2: the number of death sentences has declined since 2001 


<hr>

```{r}
sessionInfo()
```
